Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-24 paper/main.tex
edits through pdf line 345, source line 660

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. For each question, in turn, we used
Equation~\ref{eqn:prop} to estimate each participant's knowledge at the given
question's embedding space coordinate, using all \textit{other} questions that
participant answered on the same quiz. For each quiz, we grouped these
estimates into two distributions: one for the estimated knowledge at the
coordinates of \textit{correctly} answered questions, and another for the
estimated knowledge at the coordinates of \textit{incorrectly} answered
questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $\U$-tests to
compare the means of these distributions of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    Mann-Whitney $\U$-tests reported in each panel are between the distributions
    of estimated knowledge at the coordinates of correctly versus incorrectly
    answered held-out questions. In the top row (``All questions''), we used
    all quiz questions (from each quiz, for each participant) except one to
    estimate knowledge at the held-out question's embedding coordinate. In the
    middle rows (``Across-lecture''), we used all questions about one lecture
    to estimate knowledge at the embedding coordinate of a held-out question
    about the \textit{other} lecture. In the bottom row (``Within-lecture''),
    we used all but one question about one lecture to estimate knowledge at the
    embedding coordinate of a held-out question about the \textit{same}
    lecture. We repeated each of these analyses using all possible held-out
    questions for each quiz and participant.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $\U
= 50587, p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $\U = 8244, p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $\U = 8202.5, p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $\U = 7681.5, p = 0.746$; \textit{Birth of
Stars}: $\U = 8125, p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a positive shift relative
to held-out incorrectly answered questions. This held when we included all
questions in the analysis ($\U = 58332, p < 0.001$), when we predicted knowledge
across-lectures (\textit{Four Fundamental Forces}: $\U = 6749.5, p = 0.014$;
\textit{Birth of Stars}: $\U = 8480, p = 0.016$), and when we predicted
knowledge at the embedding coordinates of held-out \textit{Four Fundamental
Forces} questions using other \textit{Four Fundamental Forces} questions from
the same quiz and participant ($\U = 7224, p < 0.001$). This difference did
\textit{not} hold for within-lecture predictions of \textit{Birth of Stars}
knowledge ($\U = 7419, p = 0.739$). Again, we suggest that this might reflect a
floor effect whereby knowledge about the content of the \textit{Birth of Stars}
material is relatively low everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($\U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $\U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $\U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($\U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($\U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.

That the knowledge estimates derived from the text embedding space reliably
distinguish between held-out correctly versus incorrectly answered questions
(Fig.~\ref{fig:predictions}) suggests that the text embedding space bears at
least some relationship to participants' knowledge. But what does that
relationship look like as we move through the embedding space? For example,
suppose we know that a participant answers a question (at embedding coordinate
$X$) correctly. As we move away from $X$ in the embedding space, how does quiz
performance ``fall off'' with distance? Or, suppose the participant instead
answered that same question \textit{in}correctly. Again, as we move away from
$X$ in the embedding space, how do the chances that the participant does
\textit{not} know about the content change with distance? We reasoned that,
assuming our space is capturing something about how participants actually
organize their knowledge, conceptual knowledge right around $X$ should be
similar to the participant's knowledge of the content at $X$. And at another
extreme, at some distance (after moving sufficiently far away from $X$), our
guesses about what participants know (based on their response to the question
at location $X$) should be no better than guessing based on their overall
proportion of correctly answered questions—i.e., if $Y$ is very far away from
$X$, all we can do with the participant's response to $X$ is guess that ``their
performance on quiz questions about $Y$ is about equal to their average
performance on quiz questions about any material.''

>>>>

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. For each question, in turn, we used
Equation~\ref{eqn:prop} to predict each participant's knowledge at the given
question's embedding space coordinate, using all \textit{other} questions that
participant answered on the same quiz. For each quiz, we grouped these
predicted knowledge values into two distributions: one for the predicted knowledge at the
coordinates of \textit{correctly} answered questions, and another for the
predicted knowledge at the coordinates of \textit{incorrectly} answered
questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $\U$-tests to
compare the means of these distributions of predicted knowledge.

@@

    \caption{\textbf{Predicting knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    Mann-Whitney $\U$-tests reported in each panel are between the distributions
    of predicted knowledge at the coordinates of correctly and incorrectly
    answered held-out questions. In the top row (``All questions''), we used
    all quiz questions (from each quiz, for each participant) except one to
    predict knowledge at the held-out question's embedding coordinate. In the
    middle rows (``Across-lecture''), we used all questions about one lecture
    to predict knowledge at the embedding coordinate of a held-out question
    about the \textit{other} lecture. In the bottom row (``Within-lecture''),
    we used all but one question about one lecture to predict knowledge at the
    embedding coordinate of a held-out question about the \textit{same}
    lecture. We repeated each of these analyses using all possible held-out
    questions for each quiz and participant.}

@@

We carried out these analyses in three different ways. First, we used all (but one) of
the questions from a given quiz (and participant) to predict knowledge at the
embedding coordinate of a held-out question (``All questions'' in
Fig.~\ref{fig:predictions}). This test was intended to serve as an overall
baseline for the predictive power of our approach. Second, we used questions
about one lecture to predict knowledge at the embedding coordinate of a held-out
question about the \textit{other} lecture, from the same quiz and participant
(``Across-lecture'' in Fig.~\ref{fig:predictions}). This test was intended to
test the \textit{generalizability} of our approach by asking whether our
knowledge predictions held across the content areas of the two lectures. Third,
we used questions about one lecture to predict knowledge at the embedding
coordinate of a held-out question about the \textit{same} lecture, from the same
quiz and participant (``Within-lecture'' in Fig.~\ref{fig:predictions}). This
test was intended to test the \textit{specificity} of our approach by asking
whether our knowledge predictions could distinguish between questions about
different content covered by the same lecture. We repeated each of these
analyses using all possible held-out questions for each quiz and participant.

For the initial quizzes participants took (prior to watching either lecture),
predicted knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and predicted their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the predictions when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $\U
= 50587,~p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $\U = 8244,~p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $\U = 8202.5,~p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $\U = 7681.5,~p = 0.746$; \textit{Birth of
Stars}: $\U = 8125,~p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, predicted knowledge for
held-out questions that were answered correctly (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a significantly shift relative
to held-out questions that were answered incorrectly. This held when we included all
questions in the analysis ($\U = 58332,~p < 0.001$), when we predicted knowledge
across-lectures (\textit{Four Fundamental Forces}: $\U = 6749.5,~p = 0.014$;
\textit{Birth of Stars}: $\U = 8480,~p = 0.016$), and when we predicted
knowledge at the embedding coordinates of held-out \textit{Four Fundamental
Forces} questions using other \textit{Four Fundamental Forces} questions from
the same quiz and participant ($\U = 7224,~p < 0.001$). This difference did
\textit{not} hold for within-lecture knowledge predictions at 
knowledge at embedding space coordinates of \textit{Birth of Stars} questions ($\U = 7419,~p = 0.739$). Again, we suggest that this might reflect a
floor effect whereby knowledge about the content of the \textit{Birth of Stars}
material is relatively low everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, predicted knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($\U = 38279,~p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $\U
= 6684.5,~p = 0.032$; \textit{Birth of Stars}: $\U = 6414.5,~p = 0.002$), and
and when we carried out within-lecture knowledge predictions for held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($\U = 6126,~p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture knowledge predictions for
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($\U = 6734,~p
= 0.027$). Specifically, on Quiz 3, our knowledge predictions for held-out correctly answered questions about
\textit{Four Fundamental Forces} were reliably \textit{lower} than those for their incorrectly answered counterparts. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.  Taken together,
the results in Figure~\ref{fig:predictions} indicate that our approach can
reliably predict acquired knowledge (especially about recently learned content),
and that the knowledge predictions are generalizable across the content areas
spanned by the two lectures, while also specific enough to distinguish between
questions about more subtly different content within the same lecture.

That the knowledge predictions derived from the text embedding space reliably
distinguish between held-out correctly versus incorrectly answered questions
(Fig.~\ref{fig:predictions}) suggests that the text embedding space bears at
least some relationship to participants' knowledge. But what does that
relationship look like as we move through the embedding space? For example,
suppose we know that a participant answers a question (at embedding coordinate
$X$) correctly. As we move away from $X$ in the embedding space, how does quiz
performance ``fall off'' with distance? Or, suppose the participant instead
answered that same question \textit{in}correctly. Again, as we move away from
$X$ in the embedding space, how do the chances that the participant does
\textit{not} know about the content change with distance? We reasoned that,
assuming our space is capturing something about how participants actually
organize their knowledge, conceptual knowledge right around $X$ should be
similar to the participant's knowledge of the content at $X$. And at another
extreme, at some distance (after moving sufficiently far away from $X$), our
guesses about what participants know (based on their response to the question
at location $X$) should be no better than guessing based on their overall
proportion of correctly answered questions—i.e., if $Y$ is very far away from
$X$, all we can do with the participant's response to $X$ is guess that ``their
performance on quiz questions about $Y$ is about equal to their average
performance on quiz questions about any material.''
